# R Markdown Tutorial {#rmarkdown}

::: {.learning-goals}
ðŸŽ Learning goals: 

* Explain what Markdown is and how the syntax works, 
* Practice how to integrate code and data in RMarkdown, 
* Understand the different output formats from RMarkdown and how to generate them
* Know about generating APA format files with `papaja` and bibtex
:::

This is a short tutorial on using RMarkdown to mix prose and code together for creating reproducible scientific documents.^[The current version of the tutorial is adapted from [a slightly longer tutorial](https://github.com/mcfrank/rmarkdown-workshop) that Mike Frank and Chris Hartgerink taught together at SIPS 2017.]

In short: RMarkdown allows you to create documents that are compiled with code, producing your next scientific paper. This tutorial will help you learn the nuts and bolts of how to do this. This appendix -- actually this whole book -- is written in RMarkdown. It's a very flexible platform for writing nice looking documents.^[If you're interested in the source code for this tutorial, it's available [here](https://github.com/langcog/experimentology/blob/master/101-rmarkdown.Rmd).]

## Why write reproducible papers?

There are three reasons to write reproducible papers. To be right, to be reproducible, and to be efficient. In more depth:

1. **To avoid errors.** Using an automated method for scraping APA-formatted stats out of PDFs, @nuijten2016 found that over 10% of p-values in published papers were inconsistent with the reported details of the statistical test, and 1.6% were what they called "grossly" inconsistent, e.g. difference between the p-value and the test statistic meant that one implied statistical significance and the other did not. Nearly half of all papers had errors in them. 

2. **To promote computational reproducibility.** Computational reproducibility means that other people can take your data and get the same numbers that are in your paper. Even if you don't have errors, it can still be very hard to recover the numbers from published papers because of ambiguities in analysis. Creating a document that literally specifies where all the numbers come from in terms of code that operates over the data removes all this ambiguity.

3. **To create spiffy documents that can be revised easily.** This is actually a really big neglected one for us. At least one of us used to tweak tables and figures by hand constantly, leading to a major incentive *never to rerun analyses* because it would mean re-pasting and re-illustratoring all the numbers and figures in a paper. That's a bad thing! It means you have an incentive to be lazy and to avoid redoing your stuff. And you waste tons of time when you do. In contrast, with a reproducible document, you can just rerun with a tweak to the code. You can even specify what you want the figures and tables to look like before you're done with all the data collection (e.g., for purposes of preregistraion or a registered report). 

## Getting Started

Fire up Rstudio and create a new RMarkdown file. Don't worry about the settings, we'll get to that later.

If you click on "Knit" (or hit `CTRL+SHIFT+K`) the RMarkdown file will run and generate all results and present you with a PDF file, HTML file, or a Word file. If RStudio requests you to install packages, click yes and see whether everything works to begin with. 

We need that before we teach you more about RMarkdown. But you should feel good if you get here already, because honestly, you're about 80% of the way to being able to write basic RMarkdown files. It's _that_ easy.

## Structure of an RMarkdown file

An RMarkdown file contains several parts. Most essential are the header, the body text, and code chunks.

### Header

Headers in RMarkdown files contain some metadata about your document, which you can customize to your liking. Below is a simple example that purely states the title, author name(s), date, and output format. 

```yaml
---
title: "Untitled"
author: "NAME"
date: "July 28, 2017"
output: html_document
---
```

> ProTip: The header is written in "YAML", which means "yet another markup language." You don't need to know that, and don't worry about it. Just make sure you are careful with indenting, as YAML does care about that. 

For now, go ahead and set `html_document` to `word_document`, except if you have strong preferences for `HTML` or `PDF`.^[Note: to create PDF documents you also need a TeX installation. Don't know what that is? You probably don't have it then. More info below.]

### Body text

The body of the document is where you actually write your reports. This is primarily written in the Markdown format, which is explained in the [Markdown syntax](#markdown-syntax) section.

The beauty of RMarkdown is, however, that you can evaluate `R` code right in the text. To do this, you start inline code with \`r, type the code you want to run, and close it again with a \`. Usually, this key is below the escape (`ESC`) key  or next to the left SHIFT button.

For example, if you want to have the result of 48 times 35 in your text, you type \` r 48-35\`, which returns `r 48 - 35`. Please note that if you return a value with many decimals, it will also print these depending on your settings (for example, `r pi`).

### Code chunks

In the section above we introduced you to running code inside text, but often you need to take several steps in order to get to the result you need. And you don't want to do data cleaning in the text! This is why there are code chunks. A simple example is a code chunk loading packages. 

First, insert a code chunk by going to `Code->Insert code chunk` or by pressing `CTRL+ALT+I`. Inside this code chunk you can then type for example, `library(ggplot2)` and create an object `x`. 

```{r}
library(ggplot2)

x <- 1 + 1
```

If you do not want to have the contents of the code chunk to be put into your document, you include `echo=FALSE` at the start of the code chunk. We can now use the contents from the above code chunk to print results (e.g., $x=`r x`$).

These code chunks can contain whatever you need, including tables, and figures (which we will go into more later). Note that all code chunks regard the location of the RMarkdown as the working directory, so when you try to read in data use the relative path in.


## Markdown syntax

Markdown is one of the simplest document languages around, that is an open standard and can be converted into `.tex`, `.docx`, `.html`, `.pdf`, etc. This is the main workhorse of RMarkdown and is very powerful. You can [learn Markdown in five (!) minutes](https://learnxinyminutes.com/docs/markdown/) Other resources include [http://rmarkdown.rstudio.com/authoring_basics.html](), and [this cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf). 

You can do some pretty cool tricks with Markdown, but these are the basics:

* It's easy to get `*italic*` or `**bold**`. 
* You can get headings using `# heading1` for first level, `## heading2` for second-level, and `### heading3` for third level. Make sure you leave a space after the `#`!
* Lists are delimited with `*` for each entry.
* You can write links by writing `[here's my link](http://foo.com)`.

If you want a more extensive description of all the potential of Markdown, [this introduction to Markdown](https://daringfireball.net/projects/markdown/) is highly detailed.

The great thing about Markdown is that it works almost everywhere! Github, OSF, slack, many wikis, and even in text documents it looks pretty good. I find myself writing emails in markdown just because it's a clear and consistent way to format and outline. 

### Exercises

Swap over to your new sample markdown.

1. Outlining using headings is a really great way to keep things organized! Try making a bunch of headings, and then recompiling your document. 
2. Add a table of contents. This will involve going to the header of the document (the `YAML`), and adding some options to the `html document` bit. You want it to look like this (indentation must to be correct):

```yaml
output: 
  html_document:
    toc: true
```

Now recompile. Looks nice, right?^[Pro-tip: you can specify how deep the TOC should go by adding `toc_depth: 2` to go two levels deep]

3. Try adding another option: `toc_float: true`. Recompile -- super cool. There are plenty more great output options that you can modify. [Here is a link to the documentation.](http://rmarkdown.rstudio.com/html_document_format.html)

##Headers, Tables, and Graphs

### Headers 

We're going to want more libraries loaded (for now we're loading them inline). 

```{r}
library(knitr)
library(ggplot2)
library(broom)
library(devtools)
```

We often also add `chunk options` to each code chunk so that, for example:

- code does or doesn't display inline (`echo` setting)
- figures are shown at various sizes (`fig.width` and `fig.height` settings)
- warnings and messages are suppressed (`warning` and `message` settings)
- computations are cached (`cache` setting)

There are many others available as well. Caching can be very helpful for large files, but can also cause problems when there are external dependencies that change. An example that is useful for manuscripts is:

```{r eval=FALSE}
opts_chunk$set(fig.width=8, fig.height=5, 
               echo=TRUE, 
               warning=FALSE, message=FALSE, 
               cache=TRUE)
```


### Graphs

It's really easy to include graphs, like this one. (Using the `mtcars` dataset that comes with `ggplot2`).

```{r rmarkdown-ex}
qplot(hp, mpg, col = factor(cyl), data = mtcars)
```

All you have to do is make the plot and it will render straight into the text. 

External graphics can also be included, as follows:

```{r eval = FALSE}
knitr::include_graphics("path/to/file")
```

### Tables

There are many ways to make good-looking tables using RMarkdown, depending on your display purpose. 

- The `knitr` package (which powers RMarkdown) comes with the `kable` function. It's versatile and makes perfectly reasonable tables. It also has a `digits` argument for controlling rounding. 
- For HTML tables, there is the `DT` package, which provides `datatable` -- these are pretty and interactive javascript-based tables that you can click on and search in. Not great for static documents though. 

- For APA manuscripts, it can also be helpful to use the `xtable` package, which creates very flexible LaTeX tables. These can be tricky to get right but they are completely customizable provided you want to google around and learn a bit about tex. 

We recommend starting with `kable`:

```{r}
kable(head(mtcars), digits = 1)
```

### Statistics

It's also really easy to include statistical tests of various types. 

For this, an option is the `broom` package, which formats the outputs of various tests really nicely. Paired with knitr's `kable` you can make very simple tables in just a few lines of code. 

```{r}
mod <- lm(mpg ~ hp + cyl, data = mtcars)
kable(tidy(mod), digits = 3)
```

Of course, cleaning these up can take some work. For example, we'd need to rename a bunch of fields to make this table have the labels we wanted (e.g., to turn `hp` into `Horsepower`). 

We often need APA-formatted statistics. We can compute them first, and then print them inline.

```{r}
ts <- with(mtcars,t.test(hp[cyl==4], hp[cyl==6]))
```

> There's a statistically-significant difference in horsepower for 4- and 6-cylinder cars  ($t(`r round(ts$parameter,2)`) = `r round(ts$statistic,2)`$, $p = `r round(ts$p.value,3)`$). 

To insert these stats inline I wrote e.g. `round(ts$parameter, 2)` inside an inline code block.^[APA would require omission of the leading zero. `papaja::printp()` will let you do that, see below.]

Note that rounding can occasionally get you in trouble here, because it's very easy to have an output of $p = 0$ when in fact $p$ can never be exactly equal to 0. Nonetheless, this can help you prevent the kinds of rounding errors that would get picked up by software like `statcheck`.


## Collaboration 

How do we collaborate using RMarkdown? There are lots of different workflows that people use. The way it works in my lab is that the first author typically makes a github repository with the markdown-formatted document in it. Sometimes we just collaborate through github or through writing comments on the rendered PDF and sending them back to the first author. (I like the dropbox PDF comment interface for this). 

But, sometimes you want to do lots of line-editing or write collaboratively, especially with someone who doesn't like github and markdown and all that. For these cases, we often paste the intro into google docs or Word and edit until we converge, then the first author puts that back into the markdown. This is a little clunky, but not too bad. And critically, all the figures and numbers get rendered fresh when you re-knit, so nothing can get accidentally altered during the editing process.

## Writing APA-format papers

The end-game of reproducible research is to knit your entire paper into a submittable APA-style writeup. Managing APA format is a pain in the best of times.^[Thanks to [Frederick Aust](http://github.com/crsh) for contributing much of the code in this section!] 

### Technical requirements

We're going to use the `papaja` package. `papaja` is a R-package including a R Markdown template that can be used to produce documents that adhere to the American Psychological Association (APA) manuscript guidelines.

If you want to create PDF- in addition to DOCX-files you need **[TeX](http://de.wikipedia.org/wiki/TeX) 2013 or later**. One recent possibility is [TinyTEX](https://yihui.org/tinytex/). Or if you want a full install, try [MikTeX](http://miktex.org/) for Windows, [MacTeX](https://tug.org/mactex/) for Mac, or [TeX Live](http://www.tug.org/texlive/) for Linux.
<!-- ^[Some Linux users may need a few additional TeX packages for the LaTeX document class `apa7` to work.] -->
<!-- ^[For Ubuntu, we suggest running: `sudo apt-get install texlive texlive-publishers texlive-fonts-extra texlive-latex-extra texlive-humanities lmodern`.] -->

`papaja` has not yet been released on CRAN but you can install it from GitHub.

```{r install_papapja, eval = FALSE}
# Install devtools package if necessary
if(!"devtools" %in% rownames(installed.packages())) install.packages("devtools")

# Install papaja
devtools::install_github("crsh/papaja")
```

### Creating a document

The APA manuscript template should now be available through the RStudio menus when creating a new R Markdown file. 

When you click RStudio's *Knit* button `papaja`, `rmarkdown,` and `knitr` work together to create an APA conform manuscript that includes both your manuscript text and the results of any embedded R code.

If you don't have TeX installed on your computer, or if you would like to create a Word document replace `output: papaja::apa6_pdf` with `output: papaja::apa6_word` in the document YAML header.

`papaja` provides some rendering options that only work if you use `output: papaja::apa7_pdf`.
`figsintext` indicates whether figures and tables should be included at the end of the document---as required by APA guidelines---or rendered in the body of the document.
If `figurelist`, `tablelist`, or `footnotelist` are set to `yes` a list of figure captions, table captions, or footnotes is given following the reference section.
`lineno` indicates whether lines should be continuously numbered through out the manuscript.

### Bibiographic management

It's also possible to include references using `bibtex`, by using `@ref` syntax. An option for managing references is [bibdesk](http://bibdesk.sourceforge.net/), which integrates with google scholar.^[But many other options are possible.] 

With a bibtex file included, you can refer to papers. As an example, `@nuijten2016` results in the in text citation "@nuijten2016", or cite them parenthetically with `[@nuijten2016]` [@nuijten2016]. Take a look at the `papaja` APA example to see how this works. 

`citr` is an R package that provides an easy-to-use [RStudio addin](https://rstudio.github.io/rstudioaddins/) that facilitates inserting citations.
The addin will automatically look up the Bib(La)TeX-file(s) specified in the YAML front matter.
The references for the inserted citations are automatically added to the documents reference section. Once `citr` is installed (`install.packages("citr")`) and you have restarted your R session, the addin appears in the menus and you can define a [keyboard shortcut](https://rstudio.github.io/rstudioaddins/#keyboard-shorcuts) to call the addin.

For a bit more on `papaja`, check out [this guide](https://rpubs.com/YaRrr/papaja_guide). 


::: {.exercise}
Using the `mtcars` dataset, insert a table and a graph of your choice into your R Markdown template document. If you're feeling uninspired, try `hist(mtcars$mpg)`.
:::

::: {.exercise}
Make sure you've got `papaja`, then open a new APA template file. Compile this document, and look at how awesome it is.  Try pasting in your figure and table from your other RMarkdown (don't forget any libraries you need to make it compile). 
:::




